cardinalR: Generating interesting high-dimensional data structures

### 2025-04-21 paper-cardinalR.Rmd
Added plan for two applications
Added sample discussion..need to write after completing the application section

# Example

Add one or two datasets and evaluate how it will be useful.... (one with NLDR, one with clustering)

@@

>>>>

# Application

1) Generate a data structure with different shapes (two non-linear, one gaussian, one linear, one cone in 5-D simplex) and fit different NLDR (Then, demonstrate how this data can be used to assess the ability of different dimensionality reduction techniques (e.g., PCA, t-SNE, UMAP) to preserve the underlying structure in a lower-dimensional embedding.)

2) Add noise dimensions to the same data by linking two or three clusters, then discuss how single and complete linkage affect for the solutions (refer Di's book <https://dicook.github.io/mulgar_book/8-hierarchical.html>)

@@

The application of our high-dimensional data generation package to evaluate the interplay between dimensionality reduction, nuisance variables, and hierarchical clustering yielded several key insights. The ability to generate synthetic datasets with well-defined underlying structures, coupled with the controlled introduction of nuisance variables, provided a valuable platform for assessing the robustness of downstream unsupervised learning techniques.

Our findings demonstrated that the choice of dimensionality reduction method significantly impacted the ability of hierarchical clustering to recover the true underlying clusters. Methods that effectively preserved the global structure of the data, as defined by our generation process, generally led to more accurate and interpretable hierarchical clustering results. However, the presence of nuisance variables often confounded the low-dimensional embeddings, making it more challenging for hierarchical clustering to separate truly distinct groups. This highlights a critical consideration in real-world data analysis, where unmeasured or latent factors can obscure the signal of interest.

The hierarchical clustering analysis itself, and the choice of linkage criteria, also played a crucial role. Different linkage methods revealed varying degrees of sensitivity to the distorted representations caused by the nuisance variables. For instance, methods that prioritize compact clusters might have been more susceptible to being misled by variance introduced by nuisance factors, while others focusing on the distance between clusters might have shown more resilience. The dendrograms generated by hierarchical clustering provided a visual means to explore the relationships between samples and the potential influence of nuisance, although determining the optimal number of clusters remained a challenge in the presence of these confounding factors.

This application underscores the utility of our data generation package as a powerful tool for controlled experimentation in unsupervised learning. By providing the ground truth cluster assignments and the ability to systematically manipulate data characteristics like dimensionality, geometric structure, and the presence of nuisance variables, researchers can gain a deeper understanding of the strengths and limitations of various DR and clustering algorithms. This controlled environment allows for a more objective evaluation than often possible with real-world datasets where the underlying structure is unknown.

One limitation of this particular application was the specific type and magnitude of the nuisance variables introduced. Future work could explore a wider range of nuisance types (e.g., batch effects, technical noise with specific distributions) and their varying degrees of influence on different DR and clustering methodologies. Furthermore, investigating strategies for mitigating the impact of nuisance variables, either during the DR step or within the clustering process itself, would be a valuable extension.

In conclusion, this example demonstrates the critical role of synthetic data generation in dissecting the complex interactions within unsupervised learning pipelines. Our package provides a flexible and controlled means to create such data, enabling researchers to systematically evaluate the performance and robustness of dimensionality reduction and clustering algorithms under well-defined conditions, including the presence of confounding factors. This capability contributes to a more rigorous and informed approach to high-dimensional data analysis.

### 2025-04-21 paper-cardinalR.tex
Added sample discussion..need to write after completing the application section

@@

\section{Example}\label{example}

Add one or two datasets and evaluate how it will be useful\ldots. (one with NLDR, one with clustering)

>>>>

Add main function argument list to a table\ldots{}

@@

\section{Application}\label{application}

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\item
  Generate a data structure with different shapes (two non-linear, one gaussian, one linear, one cone in 5-D simplex) and fit different NLDR (Then, demonstrate how this data can be used to assess the ability of different dimensionality reduction techniques (e.g., PCA, t-SNE, UMAP) to preserve the underlying structure in a lower-dimensional embedding.)
\item
  Add noise dimensions to the same data by linking two or three clusters, then discuss how single and complete linkage affect for the solutions (refer Di's book \url{https://dicook.github.io/mulgar_book/8-hierarchical.html})
\end{enumerate}
